---
title: "ps5"
author: "Alex Han"
date: "2023-02-18"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

# INFO201 Problem Set: rmarkdown and plotting library(tidyverse)

## **1 Load and check data (5pt)**

1\. (1pt) For solving the problems, and answering the questions, create
a new rmarkdown docu-\
ment with an appropriate title. See
<https://faculty.washington.edu/otoomet/info201-book/>\
r-markdown.html#r-markdown-rstudio-creating.\

2\. (2pt) Load data. How many rows/columns do we have?\

```{r}
library(tidyverse) 
library(dplyr)


df <- read_delim("data/gapminder.csv")
x <- nrow(df)
y <- ncol(df)

cat("It has", x, "rows, and",y, "columns")
```

3\. (2pt) Print a small sample of data. Does it look OK?

```{r}
df %>% 
  sample_n(10)
```

The data looks OK.

## 2 Descriptive statistics (15pt)

1\. (3pt) How many countries are there in the dataset? Analyze all
three: iso3, iso2 and name.\

```{r}
count(unique(df['iso3']))
count(unique(df['iso2']))
count(unique(df['name']))
```

2\. If you did this correctly, you saw that there are more names than
iso-2 codes, and there are\
even more iso3 -codes. What is going on? Can you find it out?\

(a) (5pt) Find how many names are there for each iso-2 code. Are there
    any iso-2 codes that\
    correspond to more than one name? What are these countries?\

```{r}
df %>% 
   group_by(iso2) %>% 
   summarise(number_count = n_distinct(name)) %>%
   select(iso2,number_count) %>%
   arrange(desc(number_count)) %>%
   print()
df %>% 
   filter(is.na(name)) %>% 
   distinct(iso2)
```

The NA iso2 code correspond to more than one name.

(b) (5pt) Now repeat the same for name and iso3-code. Are there country
    names that have\
    more than one iso3-code? What are these countries?\
    Hint: two of these entitites are CHANISL and NLD CURACAO.\

```{r}
df %>% 
   group_by(name) %>% 
   summarise(name_count = n_distinct(iso3)) %>%
   select(name,name_count) %>%
   arrange(desc(name_count)) %>%
   print()
df %>% 
   filter(is.na(name)) %>% 
   distinct(iso3)
```

There are four countries with more than one iso3-code and they are:
CHANISL,GBM,KOS,NLD_CURACAO

3.  (2pt) What is the minimum and maximum year in these data?

```{r}
df %>% 
  filter(!is.na(time),!is.na(time)) %>% 
  summarise(maximum_year = max(time), minimum_year = min(time))
```

The maximum year in this data set is 2019 and the minimum year is 1960.

### 3 CO2 emissions (30pt)

*Next, let's analyze CO2 emissions.*\

1\. (2pt) How many missing co2 emissions are there for each year?
Analyze both missing CO2\
and co2_PC. Which years have most missing data?\

```{r}
df %>%
    filter(is.na(co2),is.na(co2_PC)) %>%
    group_by(time) %>% 
    summarise(years = n()) %>%
    arrange(-years) %>% 
    head(3)
```

The three years with the most missing data are: 2019, 2018, 2017

2\. (5pt) Make a plot of total CO2 emissions over time for the U.S,
China, and India. Add a few\
more countries of your choice. Explain what do you see.\

```{r}
df %>% 
    filter(name == "United States of America" | name == "China" | name == "India"
           | name == "Austria" | name == "Germany" | name == "Australia") %>%
    group_by(name, time) %>% 
    summarise(t_co2 = sum(co2)) %>% 
    ggplot(aes(x = time, y = t_co2, color = name)) +
    geom_line() +
    labs(title = "Total CO2 Emissions Over Time", x = "Year", y = "Total CO2 Emissions", color = "Country")
```

It appears to me that China had a **exponential growth** in CO2
emissions in the *early 2000'*, this might due to its **rapid economic
reforms** in the late 90's. The US is also showing a slow decline in CO2
emissions in recent years due to the *mass replacement* of *coal power*
with *solar*, *electric* and *other forms of green energy*. India is
also steadily climbing in total CO2 emissions as it is a *fast
developing* nation. The other countries I've chosen: Austria, Australia
and Germany are all pioneers in green energy and it's absolutely no
surprise that they produce substantially less CO2 emissions compared to
China, the US, and India.

3\. (5pt) Now let's analyze the CO2 emissions per capita (co2_PC ). Make
a similar plot of the\
same countries. What does this figure suggest?

```{r}
df %>% 
    filter(name == "United States of America" | name == "China" | name == "India"
           | name == "Austria" | name == "Germany" | name == "Australia") %>%
    group_by(name, time) %>% 
    summarise(t_co2pc = co2_PC) %>% 
    ggplot(aes(x = time, y = t_co2pc, color = name)) +
    geom_line() +
    labs(title = "CO2 Emissions Per Capita", x = "Year", y = "CO2 Emissions Per Capita", color = "Country")
```

It is no surprise that the US would still remain at the top of the chart
for having the highest CO2 emissions per capita. Australia, being one of
the most developed and industrialized nations of the world also situates
itself on the top of the chart. Germany and Austria while also developed
find themselves sitting in between Australia and China. The two still
developing countries of China and India have the lowest CO2 emission per
capita. This suggests that **rapidly developing** countries such as
China and India have much lower CO2 emissions per capita compared to the
developed nations. However, their total CO2 emissions are *propelled
higher* than the developed nations to due their **enormous population**.

4\. (6pt) Compute average CO2 emissions per capita across the continents
(assume region is the\
same as continent). Comment what do you see.\
Note: just compute averages over countries and ignore the fact that
countries are of different\
size.\
Hint: Americas 2016 should be 4.80.\

```{r}
df %>% 
   filter(!is.na(co2_PC)) %>%
   group_by(region) %>% 
   summarise(avg_co2_PC = mean(co2_PC)) %>% 
   print()
```

This shows that continents with more *developing nations* would produce
less CO2 per capita even if their population is much **higher** than
that of a continent with more *developed nations*.

5\. (7pt) Make a barplot where you show the previous results--average
CO2 emissions per capita\
across continents in 1960 and 2016.\
Hint: it should look something along these lines:

```{r}
ycolors <- c("royalblue4", "lightgoldenrod2")
df %>% 
  filter(time == "1960" | time == "2016") %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(!is.na(region)) %>% 
  group_by(region, time) %>% 
  summarise(avg_co2pc = mean(co2_PC)) %>% 
  ggplot(aes(x= region, y = avg_co2pc, fill = as.factor(time))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9))+
  scale_fill_manual(values = ycolors) +
  labs(x= "Continents", y = "Average CO2 Per Capita", 
       fill = "Year")
```

This shows that the average CO2 per capita has actually **dropped** for
the Americas. Personally, I believe this to be the effect of the US
switching over to *green energy* in the span of the period from 1960 to
2016. However, other continents have also **increased** in average CO2
emission per capita as the two developing *powerhouses* of Asia---
*China* and *India* are fully dedicated to the path of becoming world
economical powers. Increasing Asia's average CO2 per capita by a massive
amount from 1960 to 2016.

6\. Which countries are the three largest, and three smallest CO2
emitters (in terms of CO2 per\
capita) in 2019 for each continent? (Assume region is continent).

```{r}
#Americas
df %>% 
   filter(region == "Americas") %>% 
   filter(time == "2019") %>% 
   group_by(region) %>% 
   mutate(rank = rank(co2_PC)) %>%
   filter(rank(rank) <= 3 | rank >= n() -2) %>% 
   select(name, region, rank) %>% 
   arrange(rank)  
#Asia
df %>% 
   filter(region == "Asia") %>% 
   filter(time == "2019") %>% 
   group_by(region) %>% 
   mutate(rank = rank(co2_PC)) %>%
   filter(rank(rank) <= 3 | rank >= n() -2) %>% 
   select(name, region, rank) %>% 
   arrange(rank)  
#Europe
df %>% 
   filter(region == "Europe") %>% 
   filter(time == "2019") %>% 
   group_by(region) %>% 
   mutate(rank = rank(co2_PC)) %>%
   filter(rank(rank) <= 3 | rank >= n() -2) %>% 
   select(name, region, rank) %>% 
   arrange(rank)  
#Africa
df %>% 
   filter(region == "Africa") %>% 
   filter(time == "2019") %>% 
   group_by(region) %>% 
   mutate(rank = rank(co2_PC)) %>%
   filter(rank(rank) <= 3 | rank >= n() -2) %>% 
   select(name, region, rank) %>% 
   arrange(rank)  
#Oceania
df %>% 
   filter(region == "Oceania") %>% 
   filter(time == "2019") %>% 
   group_by(region) %>% 
   mutate(rank = rank(co2_PC)) %>%
   filter(rank(rank) <= 3 | rank >= n() -2) %>% 
   select(name, region, rank) %>% 
   arrange(rank)  
```

## 4 GDP per capita (50pt)

*Let's look at GDP per capita (GDP_PC ).*\
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by
country, using data for\
1960. Make the point size dependent on the country size, and color those
according to the\
continent. Feel free to adjust the plot in other ways to make it
better.\
Comment what do you see there.\

```{r}
con_colors <- c("darkseagreen1","khaki", "lightskyblue", "plum2", "tomato1")
df %>% 
  filter(time == "1960") %>% 
  filter(!is.na(region)) %>% 
  filter(!is.na(GDP_PC)) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region), alpha = 0.77) +
  scale_size(range = c(3, 16)) +
  scale_color_manual(values = con_colors) +
  labs(title = "GDP Per Capita Versus Life Expectancy by Country(1960", 
       x = "GDP Per Capita", y = "Life Expetancy in Years", 
       size = "Total Population", color = "Continents") +
  theme_light()
  
  
```

From the plot I'm observing that populations from countries with
**higher GDP per capita** demonstrate **higher longevity**. European
countries are often much higher on the chart in terms of GDP per capita
and life expectancy. Another observation I've made is that *population
size does not seem to have a direct effect on life expectancy* albeit
China and India are ranked relatively low in terms of life expectancy on
the chart. I believe that dots plotted toward the lower left portion of
the chart are countries yet to develop and countries plotted on the top
right portion of the graph are developed nations.

2\. (4pt) Make a similar plot, but this time use 2019 data only.\

```{r}
con_colors <- c("darkseagreen1","khaki", "lightskyblue", "plum2", "tomato1")
df %>% 
  filter(time == "2019") %>% 
  filter(!is.na(region)) %>% 
  filter(!is.na(GDP_PC)) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region), alpha = 0.77) +
  scale_size(range = c(3, 16)) +
  scale_color_manual(values = con_colors) +
  labs(title = "GDP Per Capita Versus Life Expectancy by Country(2019)", 
       x = "GDP Per Capita", y = "Life Expetancy in Years", 
       size = "Total Population", color = "Continents") +
  theme_linedraw()
```

From the graph, we can observe that countries across the world has made
**dramatic improvements** in terms of life expectancy and GDP per
capita. This is understandable as there were many **social** and
**technological advancements** made over the course of 60 years.\

3\. (6pt) Compare these two plots and comment what do you see. How has
world developed\
through the last 60 years?\

While comparing the two charts, we can observe that the world has
*developed drastically* in the span of **60 years** as more dots are
moving toward the upper right corner with countries from each continent
making progress toward development and modernization. Overall, it is
clearly that countries yet to develop in the 60's are showing great
signs of *economical* and *social* **advancements**.

4\. (6pt) Compute the average life expectancy for each continent in 1960
and 2019. Do the results\
fit with what do you see on the figures?\
Note: here as average I mean just average over countries, ignore the
fact that countries are of\
different size.

```{r}
#1960
df %>% 
   filter(!is.na(lifeExpectancy)) %>% 
   filter(grepl('1960', time)) %>% 
   group_by(region) %>% 
   summarise(avg_le60 = mean(lifeExpectancy))

#2019   
df %>% 
   filter(!is.na(lifeExpectancy)) %>% 
   filter(grepl('2019', time)) %>% 
   group_by(region) %>% 
   summarise(avg_le19 = mean(lifeExpectancy))  
```

With no surprise, life expectancy have *dramatically increased* for
across **all continents** and especially for Asia and Africa. This data
**supports** my graphs by showing an *uptrend in life expectancy* for
world populations.

5\. (8pt) Compute the average LE growth from 1960-2019 across the
continents. Show the results\
in the order of growth. Explain what do you see.\
Hint: these data (data in long form) is not the simplest to compute
growth. But you may\
want to check out the lag() function. And do not forget to group data by
continent when\
using lag(), otherwise your results will be messed up! See
<https://faculty.washington.>\
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.\

```{r}
df %>% 
   filter(!is.na(lifeExpectancy)) %>%
   filter(time == "1960"| time == "2019" ) %>% 
   filter(!is.na(region)) %>% 
   group_by(region, time) %>%
   summarise(avgle = mean(lifeExpectancy)) %>%
   mutate(prev = lag(avgle), growth = avgle - prev) %>%
   filter(!is.na(growth)) %>% 
   arrange(-growth)
```

From the output, one can observe that all continents have shown
improvements in life expectancy with the Asia and Africa demonstrating
the most substantial

improvements.

6\. (6pt) Show the histogram of GDP per capita for years of 1960 and
2019. Try to put both\
histograms on the same graph, see how well you can do it!\

```{r}
hcolors <- c("skyblue1", "sienna1")

df %>% 
  filter(time == "1960" | time == "2019") %>% 
  filter(!is.na(GDP_PC)) %>% 
  ggplot(aes(x = GDP_PC, fill = as.factor(time)))+
  geom_histogram(binwidth = 1600, position = "dodge") +
  scale_fill_manual(values = hcolors) +
  labs(title = "GDP per capita for years of 1960 and 2019", x = "GDP per Capita", fill = "Years")+
  theme_dark()

```

From the histograms, we can observe that more countries are joining the
higher end of GDP per capita in in 2019. Overall, *almost all countries*
have *shifted* over one bin to the *right* in the span of 60 years.

7\. (6pt) What was the ranking of US in terms of life expectancy in 1960
and in 2019? (When\
counting from top.)\
Hint: check out the function rank()!\
Hint2: 17 for 1960.\

```{r}
df %>% 
    filter(time == "1960") %>% 
    filter(!is.na(lifeExpectancy)) %>%
    filter(!is.na(region)) %>% 
    mutate(le_rank = rank(desc(lifeExpectancy))) %>% 
    select(name, le_rank, time,region) %>% 
    filter(name == "United States of America") %>% 
    print()
    

df %>% 
    filter(time == "2019") %>% 
    filter(!is.na(lifeExpectancy)) %>%
    filter(!is.na(region)) %>% 
    mutate(le_rank = rank(desc(lifeExpectancy))) %>% 
    select(name, le_rank, time, region) %>% 
    filter(name == "United States of America") %>% 
    print()    
   
```

The US ranked 17th in life expectancy in 1960 and 46th in 2019.

8\. (6pt) If you did this correctly, then you noticed that US ranking
has been falling quite a\
bit. But we also have more countries in 2019--what about the relative
rank divided by the\
corresponding number of countries that have LE data in the corresponding
year?\
Hint: 0.0904 for 1960.

```{r}
df %>% 
    filter(time == "1960") %>% 
    filter(!is.na(lifeExpectancy)) %>%
    filter(!is.na(region)) %>% 
    mutate(le_rank = rank(desc(lifeExpectancy)), nc = n(), r_rank = le_rank/nc) %>% 
    select(name, le_rank, time, nc, r_rank) %>% 
    filter(name == "United States of America") %>% 
    print()

df %>% 
    filter(time == "2019") %>% 
    filter(!is.na(lifeExpectancy)) %>%
    filter(!is.na(region)) %>% 
    mutate(le_rank = rank(desc(lifeExpectancy)), nc = n(), r_rank = le_rank/nc) %>% 
    select(name, le_rank, time, nc, r_rank) %>% 
    filter(name == "United States of America") %>% 
    print()
```

Using the relative rank, we can still observe that the US has dropped
significantly in terms of life expectancy over the course of 60 years.
In 1960, the US was in the 90th percentile in terms of life expectancy.
However in 2019, the US is now in the 77th percentile.

**Finally** tell us how many hours did you spend on this PS. Around 17
hours
